---
title: "NFL 2026 Season: Week 2"
# subtitle: 
description: "How historically extreme was the drop off in points scored by the Bears from week 1 to week 2?"
date: 09-26-2026
categories: [R, Chicago Bears, Points]
draft: false 
toc: false
execute:
  echo: false
  warning: false
---


```{r}
#| include: false

# for football emoji in the graph:
knitr::opts_chunk$set(dev = "ragg_png")


library(ggtext)
library(gt)
library(nflverse)
library(tidyverse)

#----------------------------------- 
########## 
#----------------------------------- 
# Custom function for NFL theme on graphs 
# Credit to Brad Congelio: https://bradcongelio.com/nfl-analytics-with-r-book/ 
  # section "1.3.2 Unique Offensive Line Combinations vs. Win Percentage": 
  
nfl_analytics_theme <- function(..., base_size = 12) {
  
  theme(
    text = element_text(family = "Calibri", #family = "Roboto",
                        size = base_size,
                        color = "black"),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16,
                                  vjust = .02,
                                  hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    strip.text = element_text(face = "bold"))
}


```


Weather certainly played a factor, and it looked like our QB Caleb Williams might have been about to score a touchdown on the play where he went down with a non-contact injury. 😖 But overall, week 2 was a struggle for my Bears and they never found the end zone. This was a dramatic change from week 1, where they scored [more points than any Bears game since 1980.](https://www.nbcsports.com/nfl/profootballtalk/rumor-mill/news/bears-panthers-was-highest-scoring-week-1-game-in-nfl-history)

**How does the huge change in points scored compare to any other week-to-week swings?**

::: {style="margin-left: 2em;"}
Nitty-gritty details: I am looking at just the regular season, and total points scored — not distinguishing whether they were offensive or defensive or special teams. I'm also going to consider either direction, negative drop offs of decreasing points or positive bounce backs of increasing points. 
:::


```{r data-team-records}
#| include: false

#----------------------------------- 
# LOAD DATA 
#----------------------------------- 
nflreadr::.clear_cache()

df_schedules <- nflreadr::load_schedules()

# Results by week
df_team_results_by_week <- bind_rows(
  # Home teams win or loss
  df_schedules %>% 
    select(season, week, game_type, 
           team = home_team, opp_team = away_team, 
           score = home_score, opp_score = away_score,
           overtime, div_game),
  # Away team win or loss
  df_schedules %>% 
    select(season, week, game_type, 
           team = away_team, opp_team = home_team,
           score = away_score, opp_score = home_score,
           overtime, div_game)
) %>% 
  mutate(result = case_when(
    score > opp_score ~ "W",
    score < opp_score ~ "L",
    score == opp_score ~ "T"
  ),
  result_for_winperc = case_when(
    score > opp_score ~ 1,
    score < opp_score ~ 0,
    score == opp_score ~ .5
  )) %>% 
  # Just regular season
  filter(game_type == "REG") %>% 
  # Exclude schedule results beyond 2026 week 2
  filter(season < 2026 | (season == 2026 & week <= 2))

# Change in score from prior week
df_team_results_by_week <- df_team_results_by_week %>% 
  # arrange for correct order for lag
  arrange(season, team, week) %>% 
  # group so we don't cross teams
  group_by(season, team) %>% 
  mutate(score_chng_weekprior = score - lag(score),
         opp_score_chng_weekprior = opp_score - lag(opp_score),
         opp_weekprior = lag(opp_team),
         weekprior = lag(week)) %>% 
  ungroup() %>% 
  relocate(c(score_chng_weekprior, 
             opp_score_chng_weekprior),
           .after = opp_score)


```

The biggest change in points scored week to week since 1999 (the earliest year available in the [nflverse](https://nflverse.nflverse.com/) data) is the Las Vegas Raiders in 2023, who went from no points in a loss to the Vikings in week 14 to putting up 63 to defeat the Chargers in week 15 (leading to [firing of the Charger's head coach Brandon Staley and general manager Tom Telesco the next day](https://www.nfl.com/news/chargers-fire-head-coach-brandon-staley-general-manager-tom-telesco-following-bl)). But that was a bounce back for the Raiders, an increase in points.

The Bears drop off from 59 points to just 3, a -56 point swing, is the biggest letdown since 1999. 

```{r}
#| out-width: "100%"

df_team_results_by_week %>% 
  slice_max(order_by = abs(score_chng_weekprior), n = 10, with_ties = FALSE) %>%
  # Teams not unique
  # mutate(team_season_week = paste0(team, " ", season, " Week ", week, " vs ", opp_team)) %>% 
  mutate(team_season_week = paste0(#team, " ", 
    season, " Week ", weekprior, " ", opp_weekprior,  " to Week ", week, " ", opp_team)) %>% 
  # Graphing
  ggplot(aes(x = score_chng_weekprior, 
             y=reorder(team_season_week, abs(score_chng_weekprior)),
             #color = team, fill = team
             )
         ) + 
  geom_col(aes(fill = ifelse(score_chng_weekprior < 0,
                             "darkred",
                             "darkgreen"))) + 
  scale_fill_identity() +
  # Add amount of change, shift inside of bar to not overlap logo
  geom_text(aes(x = score_chng_weekprior + if_else(score_chng_weekprior < 0, 11, -11),
                label = if_else(score_chng_weekprior < 0,
                                # negative has negative sign
                                as.character(score_chng_weekprior),
                                # add + to positive for emphasis
                                paste0("+", as.character(score_chng_weekprior)))
                ),
            color = "white",
            fontface = "bold",
            size = 5) +
  # own aes, no inherited color/fill
  nflplotR::geom_nfl_logos(
    aes(x = score_chng_weekprior, # 0, 
        y = reorder(team_season_week, abs(score_chng_weekprior)), 
        team_abbr = team),
    width = 0.07,
    inherit.aes = FALSE
  ) +
 scale_x_continuous(breaks = seq(-70, 70, by = 10),
                    limits = c(-70, 70)) + 
 labs(title = "Top Week-to-Week Changes in Points (1999 to Week 2 of 2026)",
       # subtitle = "",
       caption = "Regular Season only, within season (no comparison of end of season to week 1)",
       x = "Points Change vs Prior Week Played",
       y = "") +
  theme(
        panel.grid.major.y = element_blank()
        ) +
  nfl_analytics_theme() 

```
While the nflverse only has data going back to 1999, [FiveThirtyEight has a historical data set going back to 1920](https://github.com/fivethirtyeight/nfl-elo-game). Historical seasons didn't exactly have schedules with "weeks", but I will still compare each team's points to what it scored in their prior game, again limited to within-season and regular season (aka not playoffs) comparisons only.^[Minor data cleaning note: No other cleaning was done before finding the largest point changes, but prior to graphing some inconsistencies were observed and corrected with teams WSH and RED reclassified to the current abbreviation of WAS for the historical predecessors of the Washington Commanders. Note that the 1923 ARI (pictorially represented in the graph as the Arizona Cardinals) were the Chicago Cardinals, who were the historical predecessor (via St. Louis) of the Arizona Cardinals.]

```{r}
#| include: false

#----------------------------------- 
# LOAD HISTORICAL DATA 
#----------------------------------- 
# From FiveThirtyEight

df_hist_games <- read_csv('https://raw.githubusercontent.com/fivethirtyeight/nfl-elo-game/refs/heads/master/data/nfl_games.csv')


# Results by week, one row per team
# Just regular season
df_hist_team_results_by_week <- bind_rows(
  # "Home" teams win or loss (team1 is generally home team, although doesn't matter for current purposes)
  df_hist_games %>% 
    filter(playoff == 0) %>% 
      select(season, date, 
             team = team1, opp_team = team2,
             score = score1, opp_score = score2,
             neutral_538 = neutral
             ),
  # "Away" team win or loss (team2 is generally away team, although doesn't matter for current purposes)
  df_hist_games %>% 
  filter(playoff == 0) %>% 
    select(season, date, 
           team = team2, opp_team = team1,
           score = score2, opp_score = score1,
           neutral_538 = neutral
           ),
) %>% 
  mutate(result = case_when(
    score > opp_score ~ "W",
    score < opp_score ~ "L",
    score == opp_score ~ "T"
  )) %>% 
  arrange(season, team, date) %>% 
  group_by(season, team) %>% 
  # Note that doesn't match season-week as not taking into account byes
  mutate(week = row_number()) %>% 
  ungroup() 


# Change in score from prior week
df_hist_team_results_by_week <- df_hist_team_results_by_week %>% 
  # arrange for correct order for lag
  arrange(season, team, week) %>% 
  # group so we don't cross teams
  group_by(season, team) %>% 
  mutate(score_chng_weekprior = score - lag(score),
         opp_score_chng_weekprior = opp_score - lag(opp_score),
         opp_weekprior = lag(opp_team),
         weekprior = lag(week)) %>% 
  ungroup() %>% 
  relocate(c(score_chng_weekprior, 
             opp_score_chng_weekprior),
           .after = opp_score)


# historical data goes to 2020; note that week is count by team not season-week; 
  # so "weeks" in 1990 and after (when there were bye weeks) will be a bit inaccurate
  # but only comparing subsequent games, so conceptually ignoring bye weeks anyways
df_combined <- bind_rows(
  df_hist_team_results_by_week %>% mutate(source = "hist538"),
  df_team_results_by_week %>% filter(season > 2020) %>% mutate(source = "nflverse")
) %>% 
  # just some clean up
  select(source, season, week, date, 
         team, opp_team, score, opp_score, result,
         score_chng_weekprior, opp_score_chng_weekprior,
         opp_weekprior, weekprior)


```

Only three other teams have had worse point declines than the Bears.

The biggest historical decline in points in a subsequent game came in 1920 for the Rochester Jeffersons (RCH).\
They played an [informal game again a non-league military team Fort Porter (FTP), beating them 66-0 on October 10, 1920.](https://www.pro-football-reference.com/boxscores/192010100rch.htm) \
Then on [October 17, 1920 they tied 0-0 with the Utica K. of C. (UTI).](https://www.pro-football-reference.com/boxscores/192010170rch.htm) 

Only a couple other huge point swings from actual league competition were worse than the Bear's drop off: The 1973 Atlanta Falcons and the 1934 Philadelphia Eagles. That's 53 and 92 years ago, respectively.^[Update that after doing this analysis and write up, I did stumble upon [some others also digging into this.](https://x.com/benjaminsolak/status/2101774641329012796) I was focusing on total points, and not distinguishing offensive versus defensive/special team points, which is why they say they excluded the Raiders and we'll have slightly different results. I'm not sure about any of those other swings in the positive direction — if they contained a lot of defensive/special teams points, or if the other analyst was intentionally only looking at drop offs. But those details likely account for our slight inconsistencies: The [Washington 1966 game vs NYG](https://www.pro-football-reference.com/boxscores/196611270was.htm), for example, did contain defensive and special teams touchdowns.]

So the Bears didn't have the **historically worst** drop off in points *ever*, but... 😖 

```{r}
#| out-width: "100%"

### Replacing logos for historical teams
 # Could create a variable of team_logo with load_teams() links for existing teams, some other weblink for historical teams and grpah: ggimage::geom_image(aes(image = team_logo), size = ...) 

# Choose to just use football emoji
  # split the data into teams that have a logo and teams that don't, then draw each subset with its own layer
  # need this at top for emoji to show in Quarto: knitr::opts_chunk$set(dev = "ragg_png")


plot_df <- df_combined %>%
  filter(abs(score_chng_weekprior) >= 56) %>%
  mutate(
    # Manual changes, Washington; note 1923 ARI is Chicago Cardinals
    team = if_else(team == "WSH", "WAS", team),
    opp_team = if_else(opp_team == "RED", "WAS", opp_team),
    opp_weekprior = if_else(opp_weekprior == "RED", "WAS", opp_weekprior),
    # Historical "week" data is messy, so show season + matchup instead
    team_season_week = paste0(season, " ", team, " vs ", opp_weekprior, " to vs ", opp_team),
    # Fix the y-axis order once so every layer lines up
    team_season_week = fct_reorder(team_season_week, abs(score_chng_weekprior)),
    has_logo = team %in% valid_team_names(),
    bar_fill = if_else(score_chng_weekprior < 0, "darkred", "darkgreen")
  ) %>% 
  arrange(-abs(score_chng_weekprior), score_chng_weekprior)

plot_df %>%
  ggplot(aes(x = score_chng_weekprior, y = team_season_week)) +
  geom_col(aes(fill = bar_fill)) +
  scale_fill_identity() +
  # Current teams: official logos
  geom_nfl_logos(
    data = filter(plot_df, has_logo),
    aes(team_abbr = team),
    width = 0.07
  ) +
  # Historical teams: football emoji
  geom_text(
    data = filter(plot_df, !has_logo),
    label = "🏈",
    size = 6
  ) +
  # Add amount of change, shift inside of bar to not overlap logo
  geom_text(aes(x = score_chng_weekprior + if_else(score_chng_weekprior < 0, 11, -11),
                label = if_else(score_chng_weekprior < 0,
                                # negative has negative sign
                                as.character(score_chng_weekprior),
                                # add + to positive for emphasis
                                paste0("+", as.character(score_chng_weekprior)))
                ),
            color = "white",
            fontface = "bold",
            size = 5) +
  scale_x_continuous(breaks = seq(-70, 70, by = 10),
                    limits = c(-70, 70)) + 
  labs(
    title = "Top Changes in Points (1920 to Week 2 of 2026)",
    caption = stringr::str_wrap("Non-playoffs only, within season comparisons of subsequent games", width = 40),
    x = "Points Change vs Prior Game",
    y = NULL
  ) +
  nfl_analytics_theme() +
  theme(panel.grid.major.y = element_blank()
        )
```

