---
title: "NFL 2026 Season: Week 3"
# subtitle: 
description: "With a 3rd string QB, will the Bears continue their right-side running?"
date: 10-02-2026
categories: [R, Chicago Bears, Run, Pass, QB]
draft: false 
toc: false
execute:
  echo: false
  warning: false
---


```{r}
#| include: false

# knitr::opts_chunk$set(dev = "ragg_png")


library(ggtext)
library(gt)
library(nflverse)
library(tidyverse)

#----------------------------------- 
########## 
#----------------------------------- 
# Custom function for NFL theme on graphs 
# Credit to Brad Congelio: https://bradcongelio.com/nfl-analytics-with-r-book/ 
  # section "1.3.2 Unique Offensive Line Combinations vs. Win Percentage": 
  
nfl_analytics_theme <- function(..., base_size = 12) {
  
  theme(
    text = element_text(family = "Calibri", #family = "Roboto",
                        size = base_size,
                        color = "black"),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16,
                                  vjust = .02,
                                  hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    strip.text = element_text(face = "bold"))
}


# Couple of logos used on graphs later
bears_wordsmark <- 'https://github.com/nflverse/nflverse-pbp/raw/master/wordmarks/CHI.png'

bears_logo <- 'https://a.espncdn.com/i/teamlogos/nfl/500/chi.png'



```

QB1 Caleb Williams was out with the hamstring injury suffered in week 2, QB2 Tyson Bagent was in the concussion protocol for most of the week and not taking reps at practice. QB3 Case Keenum was up, and the Bears were playing a team that last year they dominated primarily by running the ball. 

Let's set the scene!

# Black Friday 2025, Week 13, Bears at Eagles

After an 0-2 start to the 2025 season, the Bears started stacking some wins, including a number of come from behind victories — but against some not-top-tier-teams (i.e., Las Vegas Raiders, New York Giants, Washington Commanders, Cincinnati Bengals). 

The game against the Eagles was a chance to make a statement [win against a quality team, and the Bears did just that, particularly with a dominating run game.](https://www.nfl.com/news/bears-eagles-on-black-friday-what-we-learned-from-chicago-s-24-15-win)

What did their run play calling look like in 2025 against the Eagles?

::: {.callout-note collapse="true"}
## Nitty-gritty details of rush attempts that are excluded

I will stick the the [definitions I used in week 1 for rush attempts](../2026-09-20-week01/#play-calling) (i.e., exclude QB scrambles and kneel downs, exclude play_type "no_play", but other plays with penalties remain).
:::

```{r data-pbp-run-load-filter}
#| include: false

#----------------------------------- 
# LOAD DATA 
#----------------------------------- 

nflreadr::.clear_cache()

# Bears v Eagles 2025 (week 13) and 2026 (week 03)
  # game_id: 2025_13_CHI_PHI, 2026_03_PHI_CHI
df_pbp_rush_chiphi_25_26 <- bind_rows(
  # 2025 regular season
  nflreadr::load_pbp(2025) %>% 
    filter(game_id %in% c("2025_13_CHI_PHI")),
  # 2026 up through week 3
  nflreadr::load_pbp(2026) %>% 
    filter(game_id %in% c("2026_03_PHI_CHI"))
)
  
# Not using run_gap_dir, just run_location
  # but leave in as filter checking below uses it to help
df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
  mutate(run_gap_dir = case_when(
      !is.na(run_gap) & !is.na(run_location) ~ paste0(run_location, "_",  run_gap),
      run_location == "middle" ~ "middle",
      TRUE ~ NA
      ) %>% factor(levels = c("left_end",
                             "left_tackle",
                             "left_guard",
                             "middle",
                             "right_guard",
                             "right_tackle",
                             "right_end"),
                  ordered = TRUE)) %>% 
  mutate(chicagopos = ifelse(posteam == "CHI", 1, 0)) %>% 
  # just rushing... may be some inconsistencies across these to check
  filter(play_type == "run" | rush_attempt == 1 | rush == 1)



#----------------------------------- 
# FILTERING 
#----------------------------------- 
# (copy from blog week 1 for consistency, but not all applicable)
  # adding # Note: None in CHI PHI 25/26 games

### exclude play_type == "no_play"
df_pbp_rush_chiphi_25_26 %>% 
  count(play_type)
df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
  filter(play_type != "no_play")


### QB Scrambles get coded as play_type == "run" and pass == 1 (which are incidentally all also rush_attempt == 1)
  # intended as passes, remove from data about rush
df_pbp_rush_chiphi_25_26 %>% 
  count(qb_scramble)
# FILTER to remove qb_scramble
df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
  filter(qb_scramble == 0)

### QB Kneel downs are not really runs
# qb_kneel account for a lot of the missing run_gap_dir
df_pbp_rush_chiphi_25_26 %>% 
  count(qb_kneel, run_gap_dir)
# FILTER to remove qb_kneel
df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
  filter(qb_kneel == 0)



### Two point conversions 
  # Note: None in CHI PHI 25/26 games
df_pbp_rush_chiphi_25_26 %>% 
  count(two_point_attempt, is.na(run_gap_dir))
# FILTER to remove two point conversions, don't have run_gap_dir
df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
  filter(two_point_attempt == 0)




### Fumbles (particularly with no gain)
df_pbp_rush_chiphi_25_26 %>% 
  count(fumble, is.na(run_gap_dir))

# Fumbles missing run_gap_dir, what is going on? 
   # break it down by whether the play gained yards 
df_pbp_rush_chiphi_25_26 %>% 
  filter(fumble == 1) %>% 
  count(is.na(run_gap_dir), yards_gained > 0)

# Breaking those down by whether it failed to gain yards: 
    # When they gain no yards, it is generally a QB fumble so no run_gap_dir because couldn't tell where they would have gone and likely never intended to be a rush
  # Note: None with no run_gap_dir in CHI PHI 25/26 games
    df_pbp_rush_chiphi_25_26 %>% 
      filter(is.na(run_gap_dir) & yards_gained <= 0 & fumble == 1) %>% 
      select(desc)
    
    # Or did gain yards:
      # Note: None in CHI PHI 25/26 games
    df_pbp_rush_chiphi_25_26 %>% 
      filter(is.na(run_gap_dir) & yards_gained > 0 & fumble == 1) %>% 
      select(desc)

# FILTER to remove when if it's a fumble AND is missing run_gap_dir
    # so keep all non-fumbles, and only keep fumbles if they have run_gap_dir
    # Note: None in CHI PHI 25/26 games
df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
  filter(fumble == 0 | (fumble == 1 & !is.na(run_gap_dir)))
    

### Illegal forward passes are not really supposed to be run plays
# Note: None in CHI PHI 25/26 games
df_pbp_rush_chiphi_25_26 %>% 
  filter(is.na(run_gap_dir)) %>% 
  select(desc)

# Other illegal forward passes
# Note: None in CHI PHI 25/26 games
df_pbp_rush_chiphi_25_26 %>% 
  filter(penalty_type == "Illegal Forward Pass") %>% 
  select(desc)

# Update: NOT removed
    # Could filter, but it's just rare; also in other years these sometimes reflect later illegal laterals on the play, not the QB; so updating to skill this removal
    # # Previously (but now not): FILTER to remove Illegal Forward Passes
# Note: None in CHI PHI 25/26 games
# df_pbp_rush_chiphi_25_26 <- df_pbp_rush_chiphi_25_26 %>% 
#   filter(is.na(penalty_type) | penalty_type != "Illegal Forward Pass") 


### Accounted for all missing on run_gap_dir
df_pbp_rush_chiphi_25_26 %>% 
  count(run_location, run_gap_dir) 


# Just penalty details for text
# Note: None in CHI PHI 25/26 games
df_pbp_rush_chiphi_25_26 %>% 
  count(penalty_team == posteam) %>% 
  mutate(prop = n / sum(n))

```


# Run Direction

In their 2025 Black Friday match up against the Eagles, the Bears were consistent with their [general trends of running to the right.](../2026-09-20-week01/index.qmd) 


```{r}
#| out-width: "100%"

# To make easier to exactly copy code for 2026 below
year <- 2025 

# keeping posteam just for ease of working with prior code for graphing
chi_playcall <-df_pbp_rush_chiphi_25_26 %>% 
  # Filter for year, CHI offense
  filter(season == year, posteam == "CHI") %>% 
  # Offensive team
  group_by(posteam, run_location) %>% 
  summarize(n = n()) %>% # still grouped by posteam
  mutate(prop = n / sum(n)) %>% ungroup() 


# Note: Altering graph code from blog week1 
  # blog week1 was top 10 team-direction combos for single team 
  # just use run_location on x, no need to create posteam_loc

# Graphing
chi_playcall %>% 
  ggplot(aes(x = run_location, 
             y = prop,
             color = posteam, fill = posteam
             )) + 
  geom_col() + 
  # value and run_location combined
  # gap between lines is controlled by the lineheight parameter 
    # (a fixed multiplier on font size, not an aesthetic — pass it outside aes())
  geom_text(aes(label = paste0(" ", round(100 * prop, 1), "%\n", 
                               str_to_title(run_location))),
            position = position_nudge(y = -0.07),
            lineheight = 0.85,
            color = "white",
            fontface = "bold",
            size = 4.5) +
  nflplotR::scale_fill_nfl(alpha = 1) +
  nflplotR::scale_color_nfl(type = "secondary") +
  # Not necessary to have logo when only one team, but logos are pretty
  # own aes, no inherited color/fill
  nflplotR::geom_nfl_logos(
    aes(x = run_location, # no need for posteam_loc
        y = prop, team_abbr = posteam),
    width = 0.08,
    inherit.aes = FALSE
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format(),
                     limits = c(0, .60),
                     expand = c(0,0)) + 
  labs(title = paste0("Bears at Eagles ", as.character(year), " Rush Directions Preferences"),
       subtitle = paste0("Black Friday ", as.character(year), " Week 13, n = ", 
       chi_playcall %>% summarize(n = sum(n)) %>% pull(n), # show how many
       " Rush Attempts"),
       x = "",
       y = "% of All Team Rush Attempts") +
  nfl_analytics_theme() + 
  theme(
      axis.text.x = element_blank(),
      panel.grid.major.x = element_blank())


```

In this 2026 week 3 Monday Night Football game, they still avoided the middle (the only run there was the 1 yard delayed QB sneak by Case Keenum for a touchdown), but they actually (slightly) preferred the left side over the right! 🤔

```{r}
#| out-width: "100%"

# NOTE: Copy of code above BUT change year 
  # AND manual title and subtitle (at or vs, Black Friday -> MNF, Week 13 -> 3)
  # OH, middle is so rare, text is hidden... need to adjust where text is for middle
year <- 2026 

# keeping posteam just for ease of working with prior code for graphing
chi_playcall <-df_pbp_rush_chiphi_25_26 %>% 
  # Filter for year, CHI offense
  filter(season == year, posteam == "CHI") %>% 
  # Offensive team
  group_by(posteam, run_location) %>% 
  summarize(n = n()) %>% # still grouped by posteam
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  # for text color alterations because middle has to go above the bar
  mutate(run_location_text_color = case_when(
    run_location == "middle" ~ "black", 
    run_location %in% c("left", "right") ~ "white"))


# Note: Altering graph code from blog week1 
  # blog week1 was top 10 team-direction combos for single team 
  # just use run_location on x, no need to create posteam_loc

# Graphing
chi_playcall %>% 
  ggplot(aes(x = run_location,
             y = prop,
             color = posteam, fill = posteam
             )) + 
  geom_col() + 
  # value and run_location combined
  # gap between lines is controlled by the lineheight parameter 
    # (a fixed multiplier on font size, not an aesthetic — pass it outside aes())
  geom_text(aes(label = paste0(" ", round(100 * prop, 1), "%\n", 
                               str_to_title(run_location)),
                # move y placement inside aes, depends on run_location
                y = prop + if_else(run_location == "middle",
                                   .07,
                                   -0.07)),
            # move into aes to change for middle: position = position_nudge(y = -0.07),
            lineheight = 0.85,
            # can't pass colors directly in aes, 
              # in aes pass variable and match labels to color with scale_color_manual
                # BUT that would override nflplotR::scale_color_nfl for bar outline
            color = chi_playcall$run_location_text_color,
            fontface = "bold",
            size = 4.5) +
  nflplotR::scale_fill_nfl(alpha = 1) +
  nflplotR::scale_color_nfl(type = "secondary") +
  # Not necessary to have logo when only one team, but logos are pretty
  # own aes, no inherited color/fill
  nflplotR::geom_nfl_logos(
    aes(x = run_location, # no need for posteam_loc
        y = prop, team_abbr = posteam),
    width = 0.08,
    inherit.aes = FALSE
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format(),
                     limits = c(0, .60),
                     expand = c(0,0)) + 
  labs(title = paste0("Bears v Eagles ", as.character(year), " Rush Directions Preferences"),
       subtitle = paste0("Monday Night Football ", as.character(year), " Week 3, n = ", 
       chi_playcall %>% summarize(n = sum(n)) %>% pull(n), # show how many
       " Rush Attempts"),
       x = "",
       y = "% of All Team Rush Attempts") +
  nfl_analytics_theme() + 
  theme(
      axis.text.x = element_blank(),
      panel.grid.major.x = element_blank())


```

What happened? During the broadcast, Troy Aikman pointed out a play where the right tackle and guard (Darnell Wright and Jonah Jackson) took off at the snap and pulled out to the left to block: "Can't always run right? Run them out to the left and then get Luther Burden in behind it."^[At the first quarter 10 minute mark in the broadcast, corresponding to a replay of nflverse play_id 271 a "left" run for a 7 yard gain.] 

::: {style="float: left; margin-right: 20px; margin-bottom: 10px; width: 326px; transform: scale(0.85); transform-origin: top left;"}

<blockquote class="instagram-media" data-instgrm-captioned data-instgrm-permalink="https://www.instagram.com/reel/Dd4H1yJxMYE/?utm_source=ig_embed&amp;utm_campaign=loading" data-instgrm-version="14" style=" background:#FFF; border:0; border-radius:3px; box-shadow:0 0 1px 0 rgba(0,0,0,0.5),0 1px 10px 0 rgba(0,0,0,0.15); margin: 1px; max-width:540px; min-width:326px; padding:0; width:99.375%; width:-webkit-calc(100% - 2px); width:calc(100% - 2px);"><div style="padding:16px;"> <a href="https://www.instagram.com/reel/Dd4H1yJxMYE/?utm_source=ig_embed&amp;utm_campaign=loading" style=" background:#FFFFFF; line-height:0; padding:0 0; text-align:center; text-decoration:none; width:100%;" target="_blank"> <div style=" display: flex; flex-direction: row; align-items: center;"> <div style="background-color: #F4F4F4; border-radius: 50%; flex-grow: 0; height: 40px; margin-right: 14px; width: 40px;"></div> <div style="display: flex; flex-direction: column; flex-grow: 1; justify-content: center;"> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; margin-bottom: 6px; width: 100px;"></div> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; width: 60px;"></div></div></div><div style="padding: 19% 0;"></div> <div style="display:block; height:50px; margin:0 auto 12px; width:50px;"><svg width="50px" height="50px" viewBox="0 0 60 60" version="1.1" xmlns="https://www.w3.org/2000/svg" xmlns:xlink="https://www.w3.org/1999/xlink"><g stroke="none" stroke-width="1" fill="none" fill-rule="evenodd"><g transform="translate(-511.000000, -20.000000)" fill="#000000"><g><path d="M556.869,30.41 C554.814,30.41 553.148,32.076 553.148,34.131 C553.148,36.186 554.814,37.852 556.869,37.852 C558.924,37.852 560.59,36.186 560.59,34.131 C560.59,32.076 558.924,30.41 556.869,30.41 M541,60.657 C535.114,60.657 530.342,55.887 530.342,50 C530.342,44.114 535.114,39.342 541,39.342 C546.887,39.342 551.658,44.114 551.658,50 C551.658,55.887 546.887,60.657 541,60.657 M541,33.886 C532.1,33.886 524.886,41.1 524.886,50 C524.886,58.899 532.1,66.113 541,66.113 C549.9,66.113 557.115,58.899 557.115,50 C557.115,41.1 549.9,33.886 541,33.886 M565.378,62.101 C565.244,65.022 564.756,66.606 564.346,67.663 C563.803,69.06 563.154,70.057 562.106,71.106 C561.058,72.155 560.06,72.803 558.662,73.347 C557.607,73.757 556.021,74.244 553.102,74.378 C549.944,74.521 548.997,74.552 541,74.552 C533.003,74.552 532.056,74.521 528.898,74.378 C525.979,74.244 524.393,73.757 523.338,73.347 C521.94,72.803 520.942,72.155 519.894,71.106 C518.846,70.057 518.197,69.06 517.654,67.663 C517.244,66.606 516.755,65.022 516.623,62.101 C516.479,58.943 516.448,57.996 516.448,50 C516.448,42.003 516.479,41.056 516.623,37.899 C516.755,34.978 517.244,33.391 517.654,32.338 C518.197,30.938 518.846,29.942 519.894,28.894 C520.942,27.846 521.94,27.196 523.338,26.654 C524.393,26.244 525.979,25.756 528.898,25.623 C532.057,25.479 533.004,25.448 541,25.448 C548.997,25.448 549.943,25.479 553.102,25.623 C556.021,25.756 557.607,26.244 558.662,26.654 C560.06,27.196 561.058,27.846 562.106,28.894 C563.154,29.942 563.803,30.938 564.346,32.338 C564.756,33.391 565.244,34.978 565.378,37.899 C565.522,41.056 565.552,42.003 565.552,50 C565.552,57.996 565.522,58.943 565.378,62.101 M570.82,37.631 C570.674,34.438 570.167,32.258 569.425,30.349 C568.659,28.377 567.633,26.702 565.965,25.035 C564.297,23.368 562.623,22.342 560.652,21.575 C558.743,20.834 556.562,20.326 553.369,20.18 C550.169,20.033 549.148,20 541,20 C532.853,20 531.831,20.033 528.631,20.18 C525.438,20.326 523.257,20.834 521.349,21.575 C519.376,22.342 517.703,23.368 516.035,25.035 C514.368,26.702 513.342,28.377 512.574,30.349 C511.834,32.258 511.326,34.438 511.181,37.631 C511.035,40.831 511,41.851 511,50 C511,58.147 511.035,59.17 511.181,62.369 C511.326,65.562 511.834,67.743 512.574,69.651 C513.342,71.625 514.368,73.296 516.035,74.965 C517.703,76.634 519.376,77.658 521.349,78.425 C523.257,79.167 525.438,79.673 528.631,79.82 C531.831,79.965 532.853,80.001 541,80.001 C549.148,80.001 550.169,79.965 553.369,79.82 C556.562,79.673 558.743,79.167 560.652,78.425 C562.623,77.658 564.297,76.634 565.965,74.965 C567.633,73.296 568.659,71.625 569.425,69.651 C570.167,67.743 570.674,65.562 570.82,62.369 C570.966,59.17 571,58.147 571,50 C571,41.851 570.966,40.831 570.82,37.631"></path></g></g></g></svg></div><div style="padding-top: 8px;"> <div style=" color:#3897f0; font-family:Arial,sans-serif; font-size:14px; font-style:normal; font-weight:550; line-height:18px;">View this post on Instagram</div></div><div style="padding: 12.5% 0;"></div> <div style="display: flex; flex-direction: row; margin-bottom: 14px; align-items: center;"><div> <div style="background-color: #F4F4F4; border-radius: 50%; height: 12.5px; width: 12.5px; transform: translateX(0px) translateY(7px);"></div> <div style="background-color: #F4F4F4; height: 12.5px; transform: rotate(-45deg) translateX(3px) translateY(1px); width: 12.5px; flex-grow: 0; margin-right: 14px; margin-left: 2px;"></div> <div style="background-color: #F4F4F4; border-radius: 50%; height: 12.5px; width: 12.5px; transform: translateX(9px) translateY(-18px);"></div></div><div style="margin-left: 8px;"> <div style=" background-color: #F4F4F4; border-radius: 50%; flex-grow: 0; height: 20px; width: 20px;"></div> <div style=" width: 0; height: 0; border-top: 2px solid transparent; border-left: 6px solid #f4f4f4; border-bottom: 2px solid transparent; transform: translateX(16px) translateY(-4px) rotate(30deg)"></div></div><div style="margin-left: auto;"> <div style=" width: 0px; border-top: 8px solid #F4F4F4; border-right: 8px solid transparent; transform: translateY(16px);"></div> <div style=" background-color: #F4F4F4; flex-grow: 0; height: 12px; width: 16px; transform: translateY(-4px);"></div> <div style=" width: 0; height: 0; border-top: 8px solid #F4F4F4; border-left: 8px solid transparent; transform: translateY(-4px) translateX(8px);"></div></div></div> <div style="display: flex; flex-direction: column; flex-grow: 1; justify-content: center; margin-bottom: 24px;"> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; margin-bottom: 6px; width: 224px;"></div> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; width: 144px;"></div></div></a><p style=" color:#c9c8cd; font-family:Arial,sans-serif; font-size:14px; line-height:17px; margin-bottom:0; margin-top:8px; overflow:hidden; padding:8px 0 7px; text-align:center; text-overflow:ellipsis; white-space:nowrap;"><a href="https://www.instagram.com/reel/Dd4H1yJxMYE/?utm_source=ig_embed&amp;utm_campaign=loading" style=" color:#c9c8cd; font-family:Arial,sans-serif; font-size:14px; font-style:normal; font-weight:normal; line-height:17px; text-decoration:none;" target="_blank">A post shared by Coach Dan Casey (@coachdancasey)</a></p></div></blockquote>
<script async src="//www.instagram.com/embed.js"></script>
:::

Even more surprising, some sharp-eyed film watchers caught a 3rd & 1 in the 4th quarter^[11:13 left in 4th quarter, 3rd & 1 Monangai ran left for a 6 yard gain, nflverse play_id 3338] where the Bears just flipped their offensive line, with right tackle and guard (Wright and Jackson) lined up on the left side. Check it out:\
⬅⬅⬅🎥 

Those two plays would make the difference of the left side vs right side slight preference in this game.

That is, the left-right split is 16 vs 15, and if we instead categorize it as literally who they choose to run behind, it flips to 17 plays in favor of Jackson and Wright's side (vs 14 plays the other direction), a 53.1% preference.

So the Bears still want to run behind Wright and Jackson, they just occasionally mixed up how they did so. 😂


```{r}
#| include: false 

# The relative success of left and right did flip between the two years 
  # Maybe a messier, complicated story... I want to get into the overall run vs pass breakdown as that was the primary story of the game 

# Success by direction 
chi_playexecution <- df_pbp_rush_chiphi_25_26 %>% 
  # Filter for year, CHI offense
  filter(posteam == "CHI") %>% 
  group_by(season, posteam, run_location) %>% 
  summarize(n = n(), # count by run_location 
            prop_success = mean(success),
            mean_yards_gained = mean(yards_gained),
            prop_1stDownorTD = mean((first_down == 1 
                               | rush_touchdown == 1)),
            .groups = "drop"
            ) %>% 
  group_by(season) %>% 
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  relocate(prop, .after = n) %>% 
  mutate(Season = factor(case_when(
    season == 2025 ~ "2025 Black Friday, Week 13",
    season == 2026 ~ "2026 MNF, Week 03")))


# Graphing
chi_playexecution %>% 
  ggplot(aes(x = run_location, 
             y = prop_success,
             fill = Season
             )) + 
  geom_col(position = "dodge") + 
  # value and run_location combined
  # gap between lines is controlled by the lineheight parameter 
    # (a fixed multiplier on font size, not an aesthetic — pass it outside aes())
  geom_text(aes(label = paste0(" ", round(100 * prop_success, 1), "%\n",
                               "Success\n", # "Rushing\n",
                               str_to_title(run_location))),
            position = position_dodge(width = .9),
            vjust = 1.3,
            #position = position_nudge(y = -0.08), #adjust here to fit
            lineheight = 0.85,
            color = "white",
            fontface = "bold",
            size = 3.5) + # adjust here to fit "Success"
  scale_fill_manual(values = c("2025 Black Friday, Week 13" = "#0B162A",
                               "2026 MNF, Week 03" = "#C83803")) +
  # nflplotR::scale_fill_nfl() +
  # nflplotR::scale_color_nfl(type = "secondary") +
  # own aes, no inherited color/fill
  # nflplotR::geom_nfl_logos(
  #   aes(x = run_location, # just run_location
  #       y = prop_success, team_abbr = posteam),
  #   width = 0.08,
  #   inherit.aes = FALSE
  # ) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format(),
                     limits = c(0, 1),
                     expand = c(0,0)
                     ) + 
  labs(title = "Bears vs Eagles Rushing Success by Direction (2025 and 2026)",
       # subtitle = "",
       x = "",
       y = "% Rushes Successful") +
  nfl_analytics_theme() + 
  theme(
      axis.text.x = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "top"
      ) 
```


If the run was dominating the game plan, we might start to dig into the success metrics. But! Despite hearing a lot of pregame talk about the strength of the Bear's run game and needing to rely on it to help a third string QB who hadn't started a regular-season game in 1,009 days... that wasn't how the game played out. 

# Play Calls: Run vs Pass

Case Keenum came out slinging the ball! I want to look at the offensive plays calls, how often the Bears called a run or pass play.

::: {.callout-note collapse="true"}
## Nitty-gritty details of classifying run and pass

I counted QB scrambles as passes because a scramble generally indicates an *plan* to pass, but a breakdown forces a run. The Bears offense only had 2 QB scrambles in 2025 and 1 in 2026 (each for 4-5 yards). I excluded QB kneel downs, punts, extra points, two point attempts (in theory although there were none), and kickoffs. Thus, the remaining offensive plays are all run or pass. 

Similar to the run plays above, I exclude any play_type of "no_play" (generally penalties before the snap or that nullify the play), but I am leaving in other plays with penalties (for now... a fuller understanding of best practices for dealing with plays with penalties is something I need to learn [as mentioned previously](../2026-09-20-week01/#play-calling)).
:::

```{r}
#| include: false

#----------------------------------- 
# LOAD DATA 
#----------------------------------- 

nflreadr::.clear_cache()

# Bears v Eagles 2025 (week 13) and 2026 (week 03)
  # game_id: 2025_13_CHI_PHI, 2026_03_PHI_CHI
df_pbp_chioff_chiphi_25_26 <- bind_rows(
  # 2025 regular season
  nflreadr::load_pbp(2025) %>% 
    filter(game_id %in% c("2025_13_CHI_PHI") &
           posteam == "CHI"),
  # 2026 up through week 3
  nflreadr::load_pbp(2026) %>% 
    filter(game_id %in% c("2026_03_PHI_CHI") &
           posteam == "CHI")
  ) 


#----------------------------------- 
# FILTERING 
#----------------------------------- 

df_pbp_chioff_chiphi_25_26 <- df_pbp_chioff_chiphi_25_26 %>% 
  filter(play_type %in% c("pass", "run")) %>% 
  # not actually any
  filter(two_point_attempt != 1)
  

# Distinguishing type of play
df_pbp_chioff_chiphi_25_26 %>% 
  count(play_type, rush, rush_attempt, 
        pass, pass_attempt, qb_scramble
        )

# 3 qb scrambles 
df_pbp_chioff_chiphi_25_26 %>% 
  filter(qb_scramble == 1) %>% relocate(desc) 


#----------------------------------- 
# SET UP 
#----------------------------------- 

# Lump scrambles in as passes
# Capitalize for graphing
df_pbp_chioff_chiphi_25_26 <- df_pbp_chioff_chiphi_25_26 %>% 
  mutate(play_call = case_when(
    play_type == "pass" ~ "Pass", # always pass and pass_attempt
    qb_scramble == 1 ~ "Pass",
    play_type == "run" & rush == 1 ~ "Run" # always rush_attempt
  )) 


```

Overall, there was a small jump in the passing rate, and a switch to preferring the pass over run — although both games were just barely on either side of being pretty evenly balanced offensive.  

```{r}
#| out-width: "100%"


df_pbp_chioff_chiphi_25_26 %>% 
  group_by(posteam, season) %>% 
  count(play_call) %>% 
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  mutate(Season = factor(case_when(
  season == 2025 ~ "2025 Black Friday, Week 13",
  season == 2026 ~ "2026 MNF, Week 03"))) %>% 
  # Visually, I want pass on the bottom
  mutate(play_call = factor(play_call, 
                            levels = c("Run", "Pass"),
                            ordered = TRUE)) %>% 
  ggplot(aes(x = Season, y = prop,
             fill = play_call,
             color = play_call)) +
  geom_col() +
  geom_text(aes(label = paste0(round(100 * prop, 1), "%\n", 
                           str_to_title(play_call))),
            position = position_stack(vjust = 0.5),
            lineheight = 0.85,
            color = "white",
            fontface = "bold",
            size = 5) +
  scale_fill_manual(values = c("Pass" = "#0B162A",
                           "Run" = "#C83803")) +
  scale_color_manual(values = c("Pass" = "#C83803",
                         "Run" = "#0B162A")) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                 labels = scales::percent_format())  +
  scale_x_discrete(position = "top")  +
  labs(title = paste0("<img src='", bears_logo,"' width='60' style='vertical-align: bottom;' />",
                  "<img src='", bears_wordsmark,"' width='120' style='vertical-align: bottom;' /> 
                  Offensive Play Calls v Eagles"),
       #subtitle = "here...",
       x = "",
       y = "% Offensive Play Calls") +
  nfl_analytics_theme() + 
  theme(
    # axis.text.x = element_blank(),
    axis.text.x = element_text(size = 12, color = "black", face = "bold"),
    panel.grid.major.x = element_blank(),
    legend.position = "none",
    # plot.title = element_markdown(face = "bold", size = 14)
  ) 


```


```{r}
#| include: false

# Just first drive?


df_pbp_chioff_chiphi_25_26 %>% 
  # First drive
  filter(drive == 1) %>%
  group_by(posteam, season) %>% 
  count(play_call) %>% 
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  mutate(Season = factor(case_when(
  season == 2025 ~ "2025 Black Friday, Week 13",
  season == 2026 ~ "2026 MNF, Week 03"))) %>% 
  # Visually, I want pass on the bottom
  mutate(play_call = factor(play_call, 
                            levels = c("Run", "Pass"),
                            ordered = TRUE)) %>% 
  ggplot(aes(x = Season, y = prop,
             fill = play_call,
             color = play_call)) +
    geom_col() +
    geom_text(aes(label = paste0(round(100 * prop, 1), "%\n",
                               str_to_title(play_call))),
              position = position_stack(vjust = 0.5),
              lineheight = 0.85,
              color = "white",
              fontface = "bold",
              size = 5) +
    scale_fill_manual(values = c("Pass" = "#0B162A",
                               "Run" = "#C83803")) +
    scale_color_manual(values = c("Pass" = "#C83803",
                             "Run" = "#0B162A")) +
    scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format())  +
  labs(title = paste0("<img src='", bears_logo,"' width='60' style='vertical-align: bottom;' />",
                      "<img src='", bears_wordsmark,"' width='120' style='vertical-align: bottom;' /> 
                      First Drive Play Calls v Eagles (2025 and 2026)"),
       #subtitle = "here...",
       x = "",
       y = "% Offensive Play Calls") +
  nfl_analytics_theme() + 
  theme(
      # axis.text.x = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "none",
      # plot.title = element_markdown(face = "bold", size = 14)
      ) 


```


But if we break it down to the four quarters, the 2025 Bears had a pass-happy 2nd quarter, but were otherwise pretty run dominant across the other three quarters. 

In 2026, we might interpret it as a 4th quarter where they ran the ball to bleed the clock and salt away the game with a commanding lead, but **the rest of the 2026 game plan was much more pass dominant, especially in the middle half of the game.** 

```{r}
#| out-width: "100%"

df_pbp_chioff_chiphi_25_26 %>% 
  # By quarters
  group_by(posteam, season, qtr) %>% 
  count(play_call) %>% 
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  mutate(Season = factor(case_when(
  season == 2025 ~ "2025 Black Friday, Week 13",
  season == 2026 ~ "2026 MNF, Week 03"))) %>% 
  # Visually, I want pass on the bottom
  mutate(play_call = factor(play_call, 
                            levels = c("Run", "Pass"),
                            ordered = TRUE)) %>% 
  mutate(qtr = case_when(
    qtr == 1 ~ "1st", qtr == 2 ~ "2nd", qtr == 3 ~ "3rd", qtr == 4 ~ "4th"
  )) %>% 
  ggplot(aes(x = qtr, y = prop,
             fill = play_call,
             color = play_call)) +
  geom_col() +
  facet_wrap(.~ Season) + 
  geom_text(aes(label = paste0(round(100 * prop, 1), "%\n",
                             str_to_title(play_call))),
            position = position_stack(vjust = 0.5),
            lineheight = 0.85,
            color = "white",
            fontface = "bold",
            size = 5) +
  scale_fill_manual(values = c("Pass" = "#0B162A",
                             "Run" = "#C83803")) +
  scale_color_manual(values = c("Pass" = "#C83803",
                           "Run" = "#0B162A")) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                   labels = scales::percent_format())  +
  labs(title = paste0("<img src='", bears_logo,"' width='60' style='vertical-align: bottom;' />",
                      "<img src='", bears_wordsmark,"' width='120' style='vertical-align: bottom;' /> 
                      Offensive Play Calls v Eagles by Quarter"),
       #subtitle = "here...",
       x = "Quarter",
       y = "% Offensive Play Calls") +
  nfl_analytics_theme() + 
  theme(
      # axis.text.x = element_blank(),
      # remove facet_wrap box behind labels:
      strip.background = element_blank(),
      strip.text = element_text(size = 12, color = "black", face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position = "none",
      # plot.title = element_markdown(face = "bold", size = 14)
      ) 


```

Ben Johnson had never coached a starting quarterback who wasn't a number 1 overall draft pick. So he decided to send Case Keenum out there throwing it all over the place. 😂 

And it was successful. 🎯

We get to some pretty small sample sizes if we were to try to look at success rates while continuing to split runs versus passes *by quarter* (i.e., in the 2026 game the Bears only had 4 runs in each of the 3rd and 4th quarters), so we'll just zoom back out to the game overall. 

The run was used less this year, but it was pretty consistent in it's success rate. But **the 2026 Bears were much more successful with pass plays against the Eagles than they were in 2025.** 


```{r}
#| include: false

# Just numbers to report 

df_pbp_chioff_chiphi_25_26 %>% 
  # By quarters
  group_by(posteam, season, qtr) %>% 
  count(play_call) %>% 
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  mutate(Season = factor(case_when(
  season == 2025 ~ "2025 Black Friday, Week 13",
  season == 2026 ~ "2026 MNF, Week 03")))
```


```{r}
#| out-width: "100%"

df_pbp_chioff_chiphi_25_26 %>% 
  group_by(posteam, season, play_call) %>% 
  summarize(n = n(), # count by play_call
            prop_success = mean(success)) %>% 
  mutate(prop = n / sum(n)) %>% ungroup() %>% 
  mutate(Season = factor(case_when(
  season == 2025 ~ "2025 Black Friday, Week 13",
  season == 2026 ~ "2026 MNF, Week 03"))) %>% 
  # Visually, I want pass on the bottom
  mutate(play_call = factor(play_call, 
                            levels = c("Run", "Pass"),
                            ordered = TRUE)) %>% 
  ggplot(aes(x = Season, y = prop_success,
             fill = play_call,
             color = play_call)) +
    geom_col(position = "dodge") +
  geom_text(aes(label = paste0(round(100 * prop_success, 1), "%\n",
                               "Success\n", # "Rushing\n",
                               str_to_title(play_call))),
            position = position_dodge(width = .9),
            vjust = 1.3,
            #position = position_nudge(y = -0.08), #adjust here to fit
            lineheight = 0.85,
            color = "white",
            fontface = "bold",
            size = 5) + # adjust here to fit "Success"
    scale_fill_manual(values = c("Pass" = "#0B162A",
                               "Run" = "#C83803")) +
    scale_color_manual(values = c("Pass" = "#C83803",
                             "Run" = "#0B162A")) +
    scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format())  +
    scale_x_discrete(position = "top")  +
  labs(title = paste0("<img src='", bears_logo,"' width='60' style='vertical-align: bottom;' />",
                      "<img src='", bears_wordsmark,"' width='120' style='vertical-align: bottom;' /> 
                      Offensive Play Success v Eagles"),
       #subtitle = "here...",
       x = "",
       y = "% Success of Play Calls") +
  nfl_analytics_theme() + 
  theme(
      axis.text.x = element_text(size = 12, color = "black", face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position = "none",
      # plot.title = element_markdown(face = "bold", size = 14)
      ) 

```
Such a fun game and a great moment for Keenum.

Bear Down! 🐻⬇️🏈

<blockquote class="instagram-media" data-instgrm-captioned data-instgrm-permalink="https://www.instagram.com/reel/Dd2xskVvmb0/?utm_source=ig_embed&amp;utm_campaign=loading" data-instgrm-version="14" style=" background:#FFF; border:0; border-radius:3px; box-shadow:0 0 1px 0 rgba(0,0,0,0.5),0 1px 10px 0 rgba(0,0,0,0.15); margin: 1px; max-width:540px; min-width:326px; padding:0; width:99.375%; width:-webkit-calc(100% - 2px); width:calc(100% - 2px);"><div style="padding:16px;"> <a href="https://www.instagram.com/reel/Dd2xskVvmb0/?utm_source=ig_embed&amp;utm_campaign=loading" style=" background:#FFFFFF; line-height:0; padding:0 0; text-align:center; text-decoration:none; width:100%;" target="_blank"> <div style=" display: flex; flex-direction: row; align-items: center;"> <div style="background-color: #F4F4F4; border-radius: 50%; flex-grow: 0; height: 40px; margin-right: 14px; width: 40px;"></div> <div style="display: flex; flex-direction: column; flex-grow: 1; justify-content: center;"> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; margin-bottom: 6px; width: 100px;"></div> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; width: 60px;"></div></div></div><div style="padding: 19% 0;"></div> <div style="display:block; height:50px; margin:0 auto 12px; width:50px;"><svg width="50px" height="50px" viewBox="0 0 60 60" version="1.1" xmlns="https://www.w3.org/2000/svg" xmlns:xlink="https://www.w3.org/1999/xlink"><g stroke="none" stroke-width="1" fill="none" fill-rule="evenodd"><g transform="translate(-511.000000, -20.000000)" fill="#000000"><g><path d="M556.869,30.41 C554.814,30.41 553.148,32.076 553.148,34.131 C553.148,36.186 554.814,37.852 556.869,37.852 C558.924,37.852 560.59,36.186 560.59,34.131 C560.59,32.076 558.924,30.41 556.869,30.41 M541,60.657 C535.114,60.657 530.342,55.887 530.342,50 C530.342,44.114 535.114,39.342 541,39.342 C546.887,39.342 551.658,44.114 551.658,50 C551.658,55.887 546.887,60.657 541,60.657 M541,33.886 C532.1,33.886 524.886,41.1 524.886,50 C524.886,58.899 532.1,66.113 541,66.113 C549.9,66.113 557.115,58.899 557.115,50 C557.115,41.1 549.9,33.886 541,33.886 M565.378,62.101 C565.244,65.022 564.756,66.606 564.346,67.663 C563.803,69.06 563.154,70.057 562.106,71.106 C561.058,72.155 560.06,72.803 558.662,73.347 C557.607,73.757 556.021,74.244 553.102,74.378 C549.944,74.521 548.997,74.552 541,74.552 C533.003,74.552 532.056,74.521 528.898,74.378 C525.979,74.244 524.393,73.757 523.338,73.347 C521.94,72.803 520.942,72.155 519.894,71.106 C518.846,70.057 518.197,69.06 517.654,67.663 C517.244,66.606 516.755,65.022 516.623,62.101 C516.479,58.943 516.448,57.996 516.448,50 C516.448,42.003 516.479,41.056 516.623,37.899 C516.755,34.978 517.244,33.391 517.654,32.338 C518.197,30.938 518.846,29.942 519.894,28.894 C520.942,27.846 521.94,27.196 523.338,26.654 C524.393,26.244 525.979,25.756 528.898,25.623 C532.057,25.479 533.004,25.448 541,25.448 C548.997,25.448 549.943,25.479 553.102,25.623 C556.021,25.756 557.607,26.244 558.662,26.654 C560.06,27.196 561.058,27.846 562.106,28.894 C563.154,29.942 563.803,30.938 564.346,32.338 C564.756,33.391 565.244,34.978 565.378,37.899 C565.522,41.056 565.552,42.003 565.552,50 C565.552,57.996 565.522,58.943 565.378,62.101 M570.82,37.631 C570.674,34.438 570.167,32.258 569.425,30.349 C568.659,28.377 567.633,26.702 565.965,25.035 C564.297,23.368 562.623,22.342 560.652,21.575 C558.743,20.834 556.562,20.326 553.369,20.18 C550.169,20.033 549.148,20 541,20 C532.853,20 531.831,20.033 528.631,20.18 C525.438,20.326 523.257,20.834 521.349,21.575 C519.376,22.342 517.703,23.368 516.035,25.035 C514.368,26.702 513.342,28.377 512.574,30.349 C511.834,32.258 511.326,34.438 511.181,37.631 C511.035,40.831 511,41.851 511,50 C511,58.147 511.035,59.17 511.181,62.369 C511.326,65.562 511.834,67.743 512.574,69.651 C513.342,71.625 514.368,73.296 516.035,74.965 C517.703,76.634 519.376,77.658 521.349,78.425 C523.257,79.167 525.438,79.673 528.631,79.82 C531.831,79.965 532.853,80.001 541,80.001 C549.148,80.001 550.169,79.965 553.369,79.82 C556.562,79.673 558.743,79.167 560.652,78.425 C562.623,77.658 564.297,76.634 565.965,74.965 C567.633,73.296 568.659,71.625 569.425,69.651 C570.167,67.743 570.674,65.562 570.82,62.369 C570.966,59.17 571,58.147 571,50 C571,41.851 570.966,40.831 570.82,37.631"></path></g></g></g></svg></div><div style="padding-top: 8px;"> <div style=" color:#3897f0; font-family:Arial,sans-serif; font-size:14px; font-style:normal; font-weight:550; line-height:18px;">View this post on Instagram</div></div><div style="padding: 12.5% 0;"></div> <div style="display: flex; flex-direction: row; margin-bottom: 14px; align-items: center;"><div> <div style="background-color: #F4F4F4; border-radius: 50%; height: 12.5px; width: 12.5px; transform: translateX(0px) translateY(7px);"></div> <div style="background-color: #F4F4F4; height: 12.5px; transform: rotate(-45deg) translateX(3px) translateY(1px); width: 12.5px; flex-grow: 0; margin-right: 14px; margin-left: 2px;"></div> <div style="background-color: #F4F4F4; border-radius: 50%; height: 12.5px; width: 12.5px; transform: translateX(9px) translateY(-18px);"></div></div><div style="margin-left: 8px;"> <div style=" background-color: #F4F4F4; border-radius: 50%; flex-grow: 0; height: 20px; width: 20px;"></div> <div style=" width: 0; height: 0; border-top: 2px solid transparent; border-left: 6px solid #f4f4f4; border-bottom: 2px solid transparent; transform: translateX(16px) translateY(-4px) rotate(30deg)"></div></div><div style="margin-left: auto;"> <div style=" width: 0px; border-top: 8px solid #F4F4F4; border-right: 8px solid transparent; transform: translateY(16px);"></div> <div style=" background-color: #F4F4F4; flex-grow: 0; height: 12px; width: 16px; transform: translateY(-4px);"></div> <div style=" width: 0; height: 0; border-top: 8px solid #F4F4F4; border-left: 8px solid transparent; transform: translateY(-4px) translateX(8px);"></div></div></div> <div style="display: flex; flex-direction: column; flex-grow: 1; justify-content: center; margin-bottom: 24px;"> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; margin-bottom: 6px; width: 224px;"></div> <div style=" background-color: #F4F4F4; border-radius: 4px; flex-grow: 0; height: 14px; width: 144px;"></div></div></a><p style=" color:#c9c8cd; font-family:Arial,sans-serif; font-size:14px; line-height:17px; margin-bottom:0; margin-top:8px; overflow:hidden; padding:8px 0 7px; text-align:center; text-overflow:ellipsis; white-space:nowrap;"><a href="https://www.instagram.com/reel/Dd2xskVvmb0/?utm_source=ig_embed&amp;utm_campaign=loading" style=" color:#c9c8cd; font-family:Arial,sans-serif; font-size:14px; font-style:normal; font-weight:normal; line-height:17px; text-decoration:none;" target="_blank">A post shared by NFL (@nfl)</a></p></div></blockquote>
<script async src="//www.instagram.com/embed.js"></script>

___
#### Data citation {#datacitation}

Carl S, Baldwin B, Sharpe L, Ho T, Edwards J (2025). *nflverse: Easily Install and Load the 'nflverse'*. R package version 1.0.3, [https://nflverse.nflverse.com/](https://nflverse.nflverse.com/).